---
title: "Prediction Assignment WriteUp"
author: "Luis Felipe"
date: "5/8/2020"
output: html_document
---



## 1 - Executive Summary

This document is the final written report of the project from the Practical Machine Learning course.  It was written and coded in RStudio, using its knitr functions and published in the html format. The purpose of this analysis is to predict the manner in which the six participants performed the exercises and answer the questions of the associated course quiz. The machine learning algorithm, which uses the classe variable in the training set, is applied to the 20 test cases available in the test data. The predictions are submitted to the Course Project Prediction Quiz for grading.

## 2 - Introduction from the Statement
Devices such as Jawbone Up, Nike FuelBand, and Fitbit can enable collecting a large amount of data about someone’s physical activity. These devices are used by the enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. However, even though these enthusiasts regularly quantify how much of a particular activity they do, they rarely quantify how well they do it. In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of six participants. They were asked to perform barbell lifts correctly and incorrectly in five different ways.

More information is available from the following website: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).


## 3 - Data Source
The data for this project can be found on the following website:

http://groupware.les.inf.puc-rio.br/har.

The training data for this project:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data for this project:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The full reference is as follows:

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. “Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human ’13)”. Stuttgart, Germany: ACM SIGCHI, 2013.

## 4 - Loading and Cleaning the Data
```{r}
#Libraries requiered are called
library(lattice)
library(ggplot2)
library(caret)
library(rpart)
library(rpart.plot)
library(corrplot)
library(rattle)
library(randomForest)
library(RColorBrewer)
set.seed(1813)
#Training and test sets are loaded
url_train <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
url_quiz  <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
data_train <- read.csv(url(url_train), strip.white = TRUE, na.strings = c("NA",""))
data_quiz  <- read.csv(url(url_quiz),  strip.white = TRUE, na.strings = c("NA",""))

```

```{r}
#Dimemensions of datasets are shown
dim(data_train); dim(data_quiz)
```

```{r}
#Two partitions are created (75 % and 25 %) within the original training dataset, and their dimmensions are shown
in_train  <- createDataPartition(data_train$classe, p=0.75, list=FALSE)
train_set <- data_train[ in_train, ]
test_set  <- data_train[-in_train, ]
dim(train_set); dim(test_set)
```
```{r}
#Since there are a lot of missing values, data is cleaned and also Since columns 1 to 5 are identification variables only, they will be removed as well.
nzv_var <- nearZeroVar(train_set)
train_set <- train_set[ , -nzv_var]
test_set  <- test_set [ , -nzv_var]
na_var <- sapply(train_set, function(x) mean(is.na(x))) > 0.95
train_set <- train_set[ , na_var == FALSE]
test_set  <- test_set [ , na_var == FALSE]
train_set <- train_set[ , -(1:5)]
test_set  <- test_set [ , -(1:5)]
dim(train_set); dim(test_set)
```

Then, the number of variables for the analysis has been reduced from the original 160 down to 54.

## 5 - Correlation Analysis

First a correlation analysis is performed between the variables before the modeling work itself is done. Select “FPC” for the first principal component order.
```{r}
corr_matrix <- cor(train_set[ , -54])
corrplot(corr_matrix, order = "FPC", method = "circle", type = "lower",
         tl.cex = 0.6, tl.col = rgb(0, 0, 0))
```

It can be seen that, if two variables are highly correlated their colors are really dark,  either dark blue (for a positive correlation) or dark red (for a negative corraltions). To further reduce the number of variables, a Principal Components Analysis (PCA) could be performed as the next step. However, since there are only very few strong correlations among the input variables, the PCA will not be performed. Instead, a few different prediction models will be built next.


## 5 - Predictive Models

Three models will be tested for the trainning data set: Decission Tree Model, GBM and Random Forest.

First Decission Tree Model is evaluated:
```{r}
set.seed(1813)
fit_decision_tree <- rpart(classe ~ ., data = train_set, method="class")
fancyRpartPlot(fit_decision_tree)
```

```{r}
#The predictions of the decision tree model on the test set are evaluated
predict_decision_tree <- predict(fit_decision_tree, newdata = test_set, type="class")
conf_matrix_decision_tree <- confusionMatrix(predict_decision_tree, test_set$classe)
conf_matrix_decision_tree
```

The predictive accuracy of the decision tree model is relatively low at 74.06 %. This can be seen in a plot

```{r}
plot(conf_matrix_decision_tree$table, col = conf_matrix_decision_tree$byClass, 
     main = paste("Decision Tree Model: Predictive Accuracy =",
                  round(conf_matrix_decision_tree$overall['Accuracy'], 4)))
```

Now the, GBM Model is evaluated:
```{r}
set.seed(1813)
ctrl_GBM <- trainControl(method = "repeatedcv", number = 5, repeats = 2)
fit_GBM  <- train(classe ~ ., data = train_set, method = "gbm",
                  trControl = ctrl_GBM, verbose = FALSE)
fit_GBM$finalModel
```

```{r}
#The the predictions of the GBM on test_set.
predict_GBM <- predict(fit_GBM, newdata = test_set)
conf_matrix_GBM <- confusionMatrix(predict_GBM, test_set$classe)
conf_matrix_GBM
```

The predictive accuracy of the GBM is relatively high at 98.88 %.

Finally the Random Forest Model is evaluated:
```{r}
set.seed(1813)
ctrl_RF <- trainControl(method = "repeatedcv", number = 5, repeats = 2)
fit_RF  <- train(classe ~ ., data = train_set, method = "rf",
                  trControl = ctrl_RF, verbose = FALSE)
fit_RF$finalModel
```

```{r}
predict_RF <- predict(fit_RF, newdata = test_set)
conf_matrix_RF <- confusionMatrix(predict_RF, test_set$classe)
conf_matrix_RF
```

The predictive accuracy of the Random Forest model is excellent at 99.92 %.

## 5 - Best Predictive Model

To summarize, the predictive accuracy of the three models evaluated is as follows:

Decision Tree Model: 74.06 %
Generalized Boosted Model: 98.88 %
Random Forest Model: 99.92 %
The Random Forest model is selected and applied to make predictions on the 20 data points from the original testing dataset (data_quiz).
```{r}
predict_quiz <- predict(fit_RF, newdata = data_quiz)
predict_quiz
```

